ODGI: scalable tools for pangenome graphs

### 2021-06-01 manuscript/main.tex
edits

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

@@

    \abstract{
        \textbf{Motivation:} Pangenome graphs provide a complete representation of
        genomic information that allows for complex population-sized analysis of genomes
        without introducing reference genome bias.
        %
        \\
        \textbf{Results:} Here we present \odgi\, a novel range of tools that makes use of an efficient
        in-memory representation of DNA graphs structured in the variation graph model. \odgi\ can
        handle hundreds of large genomes while maintaining fast execution runtime and low memory overhead.
        \odgi\ includes tools for detecting complex regions, extracting \textit{loci}, exploratory analysis,
        sorting, removing artifacts, and graphs manipulation and visualisation.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the libre MIT license. Source code and user manual can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}

@@

    \subsection{Performance on human data}

@@

    %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[!t]
      \processtable{Example of performance improvement between
        \cmd{odgi extract} and \cmd{VG something}. Centromere region
        is extracted from CHR 6 of the HPRC year 1
        assembly\citep{HPRC}, i.e., 88 haploid, phased human genome
        assemblies from 44 individuals plus the chm13 and GRCh38
        reference genomes stored in 4.3GB GFA format. All timings
        were performed on a server class machine on SSD MODEL with X
        GB of RAM and 48 CPU cores (Y x 12 core AMD Opteron) Processor
        MODEL @ 3.3 GHz with 8MB L2 Cache.\FIXME
          \label{tab:02}} {
          \resizebox{.99\columnwidth}{!}{
            \begin{tabular}{@{}lllll@{}}
                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
                \midrule
                Format conversion                           &       &       &                   \\
                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
                Subgraph extraction                         &       &       &                   \\
                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
                \botrule
            \end{tabular}}}

@@

    To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
    Thanks to the \odgi\ format's path implementation, \cmd{odgi build} embeds the paths in parallel, outperforming
    \cmd{vg convert} in the same task. For the same reason, \cmd{odgi extract} results 20 times faster than
    \cmd{vg chunk} in extracting a portion from the full graph.

>>>>

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Application Note}

@@

    \abstract{ \textbf{Motivation:} Pangenomes address the
      shortcomings of mainline genomics workflows, in particular with
      regard to reference genome bias and the identifation of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that makes use of an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes at the same time while maintaining fast
        execution runtime and low memory overheads.  \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci},
        exploratory analysis, sorting, removing artifacts, and graphs
        manipulation and visualisation.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and user manual can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu\\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}

@@

@@

    %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[!t]
      \processtable{Example of performance improvement between
        \cmd{odgi} and \cmd{VG} equivalent commands with output file
        sizes. The centromere region was extracted from CHR 6 of the
        HPRC year one assembly\citep{HPRC}, i.e., 88 haploid, phased
        human genome assemblies from 44 individuals plus the chm13
        cell line and GRCh38 reference genomes stored in 4.3GB GFA
        format.
        %
        First GFS was converted from GFA to \odgi\ and \cmd{vg}
        formats, respectively.  \odgi's path implementation can
        process paths in parallel, outperforming \cmd{vg}. For the
        same reason \cmd{odgi extract} is $20\times$ faster than
        \cmd{vg chunk} extracting a portion from the full
        graph. \FIXME: vg never finished.
        %
        All timings were performed on a server class machine
        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME
          \label{tab:02}} {
          \resizebox{.99\columnwidth}{!}{
            \begin{tabular}{@{}lllll@{}}
                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
                \midrule
                Format conversion                           &       &       &                   \\
                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
                Subgraph extraction                         &       &       &                   \\
                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
                \botrule
            \end{tabular}}}

@@
